---
title: "Analysing White Wine Quality"
author: "Parth Thakur"
date: "30 July 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE)
```

```{r, packages}

library(ggplot2)
library(gridExtra)
library(GGally)
library(psych)

```

```{r echo=FALSE, Load_the_Data}

# Loading the dataset "wineQualityWhites.csv" into a variable "df".
df <- read.csv("wineQualityWhites.csv")
df$X <- NULL 
# The variable X is the index number of the observations which is not required 
# for my analysis. Hence I removed it.
```

> In this report I am going to explore a dataset containing quality 
and other attributes of white wine. By the end of this report, I hope to 
identify the variables that impact the quality significantly.


# Univariate Plots Section

```{r echo=FALSE, Univariate_Plots}

str(df)
summary(df)
```

Our dataset contains 12 variables with more than 4000 observations. This dataset
contains no catogarical variable.

### Quality

```{r, Quality_1}

ggplot(aes(x=quality), data=df) +
    geom_histogram(binwidth=1, color="white") +
    scale_x_continuous(breaks=df$quality)
```

The quality of wine appears to be a normal distribution, with most wines with a 
quality factor of 6.

I can make the quality variable as a catagorical variable so I get seven 
levels of quality for our wines.

```{r, Quality_2}

df$quality <- factor(df$quality)
summary(df$quality)
```

Now I can clearly see that there are only 5 wines with a quality factor of 9. 
I wonder which variable affects the quality of the wine the most?

### Density

```{r, Density_1}

ggplot(aes(x=density), data=df) +
    geom_histogram()
```

```{r, Density_2}
ggplot(aes(x=density), data=df) +
    geom_histogram(bins=200) +
    xlim(0.985, 1)

summary(df$density)
```

The density of the wine doesn't appear to vary much. With interquartile range of
approximately 0.004, it is fair to assume that the density of the wine does not 
affect the quality.


### Alcohol

```{r, Alcohol_1}

ggplot(aes(x=alcohol), data=df) +
    geom_histogram()
```

```{r, Alcohol_2}
ggplot(aes(x=alcohol), data=df) +
    geom_histogram(bins=100) 
summary(df$alcohol)
```

The distribution alcohol content in the wine appears to be normal. 
Does higher alcohol content translates to better quality of wine?


### Acidity

```{r, Acidity_1}

a1 <- ggplot(aes(x=fixed.acidity), data=df) +
    geom_histogram(bins=100)
a2 <- ggplot(aes(x=volatile.acidity), data=df) +
    geom_histogram(bins=100)
a3 <- ggplot(aes(x=citric.acid), data=df) +
    geom_histogram(bins=100)

grid.arrange(a1, a2, a3, ncol=1)
```

```{r, Acidity_2}

a4 <- a1  +  scale_x_log10()
a5 <- a2  +  scale_x_log10()
a6 <- a3  +  scale_x_log10()

grid.arrange(a4, a5, a6, ncol=1)
```

The log transformed plots of the acidity attributes reveal a normal distribution
. With volatile acidity having the highest spread, and citric acidity having the 
lowest, which one of these attributes impact quality the most? It is important 
to note that the values on the x-axis are in gram/Litre and not pH

```{r, pH_1}

ggplot(aes(x=pH), data=df) +
    geom_histogram(bins=100) +
    scale_x_continuous(breaks=(seq(2.7, 3.7, by=0.1)))
summary(df$pH)
```

Most wines have pH between 2.9 and 3.50.
The median is 3.18 and the max value is 3.820.
The pH of a substance can add to the substance's "sourness" or "bitterness". A 
high pH corresponds to bitter wines, while lower pH attributes to sour wine.


### Residual Sugar

```{r, Sugar_1}

ggplot(aes(x=residual.sugar), data=df) +
    geom_histogram(bins=100)
```


```{r, Sugar_2}

ggplot(aes(x=residual.sugar), data=df) +
    geom_histogram(bins=100) +
    scale_x_log10(breaks=seq(1, 10))
```

Transofrmed to take a better look at the distribution of residual sugar, the 
data appears to be bimodal with residual sugar peaking first at 1.5 and then 
again at 10.

Residual sugar refers to the sugar that remains in the wine after fermentation 
stops. Higher residual sugar would make the wine sweet. Are sweet wines of 
better quality?


### Chlorides

```{r, Chlorides_1}

ggplot(aes(x=chlorides), data=df) +
    geom_histogram(bins=100)
```


```{r, Chlorides_2}

ggplot(aes(x=chlorides), data=df) +
    geom_histogram(bins=100) +
    scale_x_log10(breaks=c(0.01, 0.025, 0.065, 0.1))
summary(df$chlorides)
```

Chlorides in wines have a normal distribution with most wines having chlorides 
between 0.025 and 0.065.


### Sulfur Dioxide (SO2)

```{r, S02_1}

SO2_free <- ggplot(aes(x=free.sulfur.dioxide), data=df) +
    geom_histogram(bins=100) +
    scale_x_continuous(breaks=c(0, 30, 100, 200, 300))

SO2_total <- ggplot(aes(x=total.sulfur.dioxide), data=df) +
    geom_histogram(bins=100) +
    scale_x_continuous(breaks=c(0, 100, 120, 200, 300, 400))

Sulphates <- ggplot(aes(x=sulphates), data=df) +
    geom_histogram(bins=100)

grid.arrange(SO2_free, SO2_total, Sulphates, ncol=1)
```

Free SO2 and total SO2 have a normal distribution. With free SO2 peaking at 
about 30, and total SO2 peaking at about 120.
Sulphates also have a normal distribution.


# Univariate Analysis

### What is the structure of your dataset?

There are 4898 observations with 13 features. (fixed.acidity, volatile.acitidy,
citric.acid, residual.sugar, chlorides, free.sulfur.dioxide, 
total.sulfur.dioxide, pH, sulphates, alcohol, quality). The feature, quality, is
a  ordered factor variable with the following levels.

(Worst) ------------> (Best) <br>
**quality:** 3, 4, 5, 6, 7, 8, 9

Other observations: <br>
1. Most wines have a quality factor of 6.
2. The inter-quantile range of density of wines is 0.0044.


### What is/are the main feature(s) of interest in your dataset?

The main feature of interest is the quality of the wines. I'd like to determine 
which factors affect the quality of a wine the most. I suspect residual sugar 
and some other combination of variables could be used to build a predictive 
model for the quality of wines.

### What other features in the dataset do you think will help support your
investigation into your feature(s) of interest?

Alcohol content, acidity, and residual sugar are likely to contribute to the 
quality of a wine. I think acidity and residual sugar contribute the most to 
quality based on anecdotal evidence.

### Did you create any new variables from existing variables in the dataset?

I did not create any new variable in the dataset, but I converted the quality 
variable to a catagorical factor.

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

I did not find any unusual data or change any variable in the dataset.


# Bivariate Plots Section

```{r, out.width="120%", Bivariate_Plots}

pairs.panels(df, cex.cor=1.5, gap=0.2, pch=".")
```

From the plot above, quality is the most correlated to alcohol. There is also a 
strong negative correlation between density and quality. In the previous section
, I had assumed that density wouldn't affect the quality of the wine. Maybe that 
assumptions was wrong. I need to take a closer look at the scatterplots.


```{r, Quality-Density_1}

ggplot(aes(x=quality, y=density), data=df) +
    geom_boxplot() +
    scale_x_discrete(breaks=df$quality)
```

```{r, Quality_Density_2}

ggplot(aes(x=quality, y=density), data=df) +
    geom_boxplot(alpha=0.1) +
    scale_y_continuous(limits=c(0.985, 1))
```

Compairing density to quality, the first plot suffers from overplotting. zooming
in by changing the limits on the axes, the corellation is still not too 
apparent. Perhaps some other factors affect the density in such a way that it 
correlates to quality.


```{r, Quality-Alcohol_1}
ggplot(aes(y=alcohol, x=quality), data=df) +
    geom_boxplot()
```

Here I observe a very strong correlation between alcohol content and quality. 
Let me investigate further and check weather I get some other strong 
correlations.

```{r, Quality-Resibual.Sugar_1}

ggplot(aes(x=quality, y=residual.sugar), data=df) +
    geom_boxplot()
```

```{r, , Quality-Resibual.Sugar_2}
ggplot(aes(x=quality, y=residual.sugar), data=df) +
    geom_boxplot(alpha=0.1) +
    scale_y_continuous(limits=c(0, 25))
```

The above plots of Quality versus residual sugar does not give us a lot of 
insight for the quality of the wine. I will continue exploring other variables.

```{r, Quality-Others_1}

qo1 <- ggplot(aes(y=chlorides, x=quality), data=df) +
        geom_boxplot() +
        coord_cartesian(ylim=c(0, 0.065))

qo2 <- ggplot(aes(y=pH, x=quality), data=df) +
        geom_boxplot() +
        coord_cartesian(ylim=c(2.8, 3.4))

qo3 <- ggplot(aes(y=fixed.acidity, x=quality), data=df) +
        geom_boxplot(alpha=0.1) +
        coord_cartesian(ylim=c(5, 9))

qo4 <- ggplot(aes(y=total.sulfur.dioxide, x=quality), data=df) +
        geom_boxplot(alpha=0.1) +
        coord_cartesian(ylim=c(50, 250))

grid.arrange(qo1, qo2, ncol=2)
```

From the plots above I can see a positive corellation between pH and quality, 
and a negative correlation between chlorides and quality.

```{r, Quality-Others_2}
grid.arrange(qo3, qo4, ncol=2)
```

The fixed acidity and total sulfur dioxide does not seem to affect quality much.

I will now investigate relationships between density variables to get an insight 
into the high correlation between Density and Quality..

```{r, Residual.sugar-Density_1}

ggplot(aes(y=density, x=residual.sugar), data=df) +
    geom_point()
```

```{r, Residual.sugar-Density_2}

ggplot(aes(y=density, x=residual.sugar), data=df) +
    geom_point(alpha=0.5) +
    coord_cartesian(ylim=c(0.985, 1.005),
                    xlim=c(0, 30))
```

The plot between density and residual sugar suffers from over plotting. 
Adding transparency and changing the axis limits, I can see a strong 
correlation between the two variables. I wonder how this plot will look like if 
I colour the points with quality.


```{r, Alcohol-Residual.Sugar_1}

ggplot(aes(x=alcohol, y=residual.sugar), data=df) +
    geom_jitter(alpha=0.1) +
    coord_cartesian(ylim=c(0, quantile(df$residual.sugar, 0.99)))
```

Residual sugar also seems to be correlated with alcohol content. Let me chack 
how alcohol content is related to density.

```{r, Density-Alcohol}

ggplot(aes(y=alcohol, x=density), data=df) +
    geom_point(alpha=0.1) +
    coord_cartesian(xlim=c(quantile(df$density, 0.001), 
                           quantile(df$density, 0.99)))
```

Alcohol and density are also strongly correlated. I can clearly see that as 
density increases, alcohol content decreases.

```{r, Total.SO2-Density_1}

ggplot(aes(y=density, x=total.sulfur.dioxide), data=df) +
    geom_point(alpha=0.1) +
    coord_cartesian(xlim=c(0, quantile(df$total.sulfur.dioxide, 0.99)),
                    ylim=c(quantile(df$density, 0.001), 
                           quantile(df$density, 0.99))) 
```

The total surphur dioxide in the wine also seem to influence density a lot.


# Bivariate Analysis


### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

In this section I observed that quality is correlated to density the most. There
is also a strong correlation between quality and alcohol, chlorides, and pH. 
Other variables such as residual sugar and total sulfur dioxide do not seem to 
affect the quality of the wine directly, but they influence density.

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

The most interesting relationship was between residual sugar and alcohol. The 
appear to be negatively correlated, which makes sense as the fermentation 
process by which wines are made, convert sugar into alcohol.

Other interesting relationship was between alcohol and density. The density of 
the wine was minimum when the alcohol content was higher. 

### What was the strongest relationship you found?

The strongest relationship was found between density and residual sugar. Density
of the wine almost entirely depends on the amount of sugar in the wine. This 
also explains the strong correlation between density and quality.

# Multivariate Plots Section


```{r, Multivariate_Plots}

mp1 <- ggplot(aes(x=residual.sugar, y=density, color=quality), data=df) +
        geom_point(alpha=0.5) +
        coord_cartesian(xlim=c(0, quantile(df$residual.sugar, 0.99)),
                        ylim=c(quantile(df$density, 0.01), 
                               quantile(df$density, 0.99))) +
        guides(colour = guide_legend(override.aes = list(alpha=1)))

mp2 <- ggplot(aes(x=alcohol, y=density, color=quality, fill=quality), data=df) +
        geom_point(alpha=.5) +
        coord_cartesian(xlim=c(quantile(df$alcohol, 0.01), 
                               quantile(df$alcohol, 0.99)),
                        ylim=c(quantile(df$density, 0.01), 
                               quantile(df$density, 0.99))) +
        guides(colour = guide_legend(override.aes = list(alpha=1)))

grid.arrange(mp1, mp2, ncol=2)
```

In the plots above it is getting difficult to see the relationship between the 
variables. I'd like to create some catagorical variable buckets from the 
existing variables to make visualising data easier.


```{r, Make new variables}

density.breaks <- c(0.98, 0.992, 0.994, 0.996, 1.1)
sugar.breaks <- c(0, 3, 66)
sglb <- c("Low", "High")
alcohol.breaks <- c(0, 9.5, 10.4, 11.4, 15)
chlorides.breaks <- c(0, 0.036, 0.043, 0.05, 0.35)
quality.breaks <- c(0, 3, 4, 10)
quality.labels <- c("poor", "average", "great")

```

```{r echo=TRUE, "Show code"}
df$density.bucket <- cut(df$density, breaks=density.breaks)
df$sugar.bucket <- cut(df$residual.sugar, breaks=sugar.breaks, labels=sglb)
df$alcohol.bucket <- cut(df$alcohol, breaks=alcohol.breaks)
df$chlorides.bucket <- cut(df$chlorides, breaks=chlorides.breaks)
df$quality.bucket <- cut(as.numeric(df$quality), breaks=quality.breaks, 
                         labels=quality.labels)
```

```{r, show new variables}

str(df)
```

Now I have four more catagorical variables as seen above. I split alcohol, 
density, and chlorides using the quantile values, while I split residual sugar 
in two buckets to emphasise it's bimodat distribution.
I also greated a quality bucket. <br>
quality < 6 : Poor <br>
quality = 6 : Average <br>
quality 7 and above : Great

Let me use these new variables in my analysis.

```{r, residual.sugar-alcohol-density_1}

ggplot(aes(x=alcohol, y=density, color=sugar.bucket), data=df) +
    geom_jitter(width=0.5, alpha=0.5) +
    coord_cartesian(ylim=c(quantile(df$density, 0.01), 
                           quantile(df$density, 0.99)))
```

Quality was very strongly correlated to density of the wine. In the above plot, 
it is clear that density and alcohol content are very much dependent on the 
residual sugar.

I want to use multivariate analysis to get an insight in how these variables 
react with each other to influence quality.

```{r, residual.sugar-alcohol-density-quality}

ggplot(aes(y=density, x=alcohol, color=sugar.bucket), data=df) +
    geom_jitter(alpha=0.5) +
    facet_wrap(~quality.bucket) +
    coord_cartesian(ylim=c(quantile(df$density, 0.01), 
                           quantile(df$density, 0.99)))
```

Facet wrapping the plot by quality, I can see that the median of density in 
white wines decreases with quality.

```{r, quality-density-alcohol}

ggplot(aes(x=quality.bucket, y=density, color=alcohol.bucket), data=df) +
    geom_jitter(alpha=0.5) +
    coord_cartesian(ylim=c(quantile(df$density, 0.01), 
                           quantile(df$density, 0.99)))
```

```{r, quality-density-alcohol-sugar}

ggplot(aes(x=quality.bucket, y=density, color=alcohol.bucket), data=df) +
    geom_jitter(alpha=0.5) +
    coord_cartesian(ylim=c(quantile(df$density, 0.01), 
                           quantile(df$density, 0.99))) +
    facet_wrap(~sugar.bucket)
```

The first plot confirms my belief that Alcohol content and Density greatly 
affects the quality of white wine. Facet wrapping the plot by residual sugar, 
the amount of influence of residual sugar on the wine quality is clear.

I think I now have a clear understanding of what affects the quality of wines 
the most. But before reaching a definate conclusion, I'd like to check if I am 
missing correlations with other variables.

```{r, quality-others}

ggplot(aes(x=fixed.acidity, y=volatile.acidity), data=df) +
    geom_jitter(alpha=0.5) +
    facet_wrap(~quality.bucket)

ggplot(aes(x=pH, y=chlorides), data=df) +
    geom_jitter(alpha=0.5) +
    facet_wrap(~quality.bucket) +
    coord_cartesian(ylim=c(0, quantile(df$chlorides, 0.99)))

ggplot(aes(x=sulphates, y=total.sulfur.dioxide), data=df) +
    geom_jitter(alpha=0.5) +
    facet_wrap(~quality.bucket)
```

I facet wrapped the above plots by quality. This gave me a quick way to review 
the relationships.

The three charts above have some interesting relationships which may affect the 
quality of white wine. For example, from the first chart, the variance in 
volatile acidity decreases with increasing quality. The same is true for total 
sulfur dioxide. Unfortunately, these plots are not at noteworthy as some other 
relationships found, and a confident model cannot be created using these 
features.


I think, I now have a pretty good understanding of which features influence the 
quality of the wine the most. With a very strong correlation, these features are
of most interest: <br>
1. Alcohol Content <br>
2. Density <br>
3. Residual Sugar <br>

There are other features that may influence the quality, but there isn't enough
data to reach a conclusion. Some of these features are, <br>
1. Chlorides <br>
2. Total Sulfur dioxide <br>
3. Sulphates <br>
4. pH <br>

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

In this part of my investigation I was able to strengthen my beliefs about the 
varibles that influenced quality. I was able to get insights into density, 
alcohol content, and residual sugar, which had the most significant correlation 
with quality.

I was able to get a deeper insight between those variables, mainly that the 
relationship between density, alcohol, and residual sugar remains consistent 
with quality. Some other variables like volatile acidity, and total sulfur 
dioxide had some correlation with quality, but the relationships didnot seem 
strong enough.

### Were there any interesting or surprising interactions between features?

In the univariate section of this report, I assumed that density wouldn't affect
quality much. But I quickly learned that this assumption was incorrect. Density 
is highly correlated with alcohol content, and from the very begining, it was 
apparent that wines with higher alcohol content were of better quality. This was
surprising because alcohol itself doesn't have any taste.

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

I did not create any linear model as I feel there isn't enough data to make a 
confident model. There are only five observations of highest quality wine, which
is too little to determine a relationship.

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}

p1_1 <- ggplot(aes(x=quality, y=density), data=df) +
        geom_boxplot() +
        ylim(c(quantile(df$density, 0.01), quantile(df$density, 0.99))) +
        ggtitle("Relationship between Quality and Density.") +
        ylab("Density (g/cubic cm)") +
        xlab("")

p1_2 <- ggplot(aes(x=quality, y=alcohol), data=df) +
        geom_boxplot() +
        ggtitle("Relationship between Quality and Alcohol.") +
        ylab("Alcohol (% by volume)") +
        xlab("Quality (Higher is better)")

grid.arrange(p1_1, p1_2, ncol=1)
```

### Description One

This plot represents a key finding in my report. That quality of white wine is 
significantly correlated to it's density and it's alcohol content. These were 
the two of the strongest correlations to the quality. Density at -0.31 and 
alcohol at 0.44. <br>

In the box plot at the top, the downward trend in the density of the wine tell 
that quality of wine increases with decrease in the density. I changed the 
limits on the y-axis to exclude the bottom 1% and the top 1% of the data to make
it easier to see this trend. In the second boxplot, I see the relationship 
between alcohol content and quality. I can see an upeard trend, however there 
is a dip in alcohol content before it starts to rise again. <br>

This dip makes me believe that there is a certain threshold for alcohol percent 
before it starts to have a positive impact on quality. In fact I am certain, 
that if alcohol content keeps on rising, quality will start to dip again as 
higher concentrations of alcohol impart a bitter taste to wines. <br>

### Plot Two
```{r echo=FALSE, Plot_Two}

ggplot(aes(x=alcohol, y=density, color=sugar.bucket), data=df) +
    geom_point(alpha=0.5) +
    ylim(c(quantile(df$density, 0.01), quantile(df$density, 0.99))) +
    ggtitle("Relationship between the three most important variables.") +
    labs(x="Alcohol content (% by volume)",
         y="Density (g/cubic cm)",
         color="Residual sugar")
```

### Description Two

In this plot I see the relationship between the three most significant 
variables. Specifically, repltionship between density and alcohol, and residual 
sugar. I split the sugar values in two to better visualize the data. Wines with 
residual sugar less than 3g/dm^3 are "Low sugar", and those with more than that 
are "High sugar". <br>

These three variables show the highest amount of correlation. As alcohol content
increases, residual sugar and density decreases. This makes sense, as during the
fermentation process, sugar is converted to alcohol. Since sugar is a heavy 
compound and alcohol is very light, the density goes on decreasing. <br>

This relationship also helps us understand the age of the wine. The older the
wine, the longer it is fermented, which gives it a high alcohol content and a 
lower density. Hence I can also say that older wines are of better quality. <br>

### Plot Three
```{r echo=FALSE, Plot_Three}

ggplot(aes(x=sugar.bucket, y=density, color=alcohol.bucket), data=df) +
    geom_jitter(alpha=0.5) +
    coord_cartesian(ylim=c(quantile(df$density, 0.01), 
                           quantile(df$density, 0.99))) +
    facet_wrap(~quality.bucket) +
    labs(y="Density (g/cubic cm)",
         x="Residual Sugar",
         color="Alcohol content (% by volume)") +
    scale_color_discrete(labels=c("8 - 9.5", "9.5 - 10.4", 
                                  "10.4 - 11.4", "11.4 - 14.2"))
```

### Description Three

This plot shows the relationship between all our important variables. I have 
facet wrapped the plot based on the quality of the wine, and added some jitter 
to the scatter plot to make it easier to visualize the date.

The relationship between residual sugar and density is very clear in the plot. 
With all high sugar wines having a higher median density than low sugar wines. 
By looking at the colour distribution of the points, I can also observe that 
great quality alcohols have the highest concentration of high alcohol wines, 
while poor quality alcohols have the highest concentration of high sugar-low 
alcohol wines.

Hence I can conclude that sweet wines are in fact not of great quality, and 
that the quality of wines is determined mostly by its alcohol content, and by 
extention its density and age.

------

# Reflection


To say this project was challenging would be an understatement. I had to face a 
lot of obstacles while completing this project, which only made my interest in 
data analytics grow. To put simply, this project taught me how to look at data 
and how to translate it to visualizations to better understand it. I became more
familier with the various types of plots, like scatter plots, box plots, 
histograms, and even though I did not make use of them in this project, line 
plots. <br>

The univariate and bivariate section was simple to do. I only had to look at the
data distribution, and it's relationships with other important variables. The 
multivariate section though, proved to be challenging. Even though I had the 
idea of what my conclusion would be, it was difficult to express it by way of 
plots. The dataset only had one factor variable, and it was getting difficult to
make simple plots with just the quality. <a 
href=https://github.com/dwolfATL/White-Wine-EDA/blob/master/EDA-whitewine.html>
Daniel Wolf's white wine exploration </a> gave me the idea of splitting some 
variables like sugar and alcohol into buckets and making new catagorical 
variables. <br>

This project made me realise that real world data will not always have the 
variables I need, and that it is important to be creative and make new variables
as and when required.

<!---
References:

[1] Diamonds Exploration by Chris Saden found at
https://s3.amazonaws.com/content.udacity-data.com/courses/ud651/diamondsExample_2016-05.html

[2] White Wine Exploratory Data Analysis by Daniel Wolf found at
https://github.com/dwolfATL/White-Wine-EDA/blob/master/EDA-whitewine.html
---!>

